Types from data: \textnormal{Making structured data first-class citizens in F\#}

### 2015-11-19 pldi-2016/paper.tex
Final edits, page 1

\title{F\# Data: \textnormal{Making structured data first-class citizens}}
%\subtitle{Subtitle Text, if any}

@@

In this paper, we integrate external structured data into the F\# type system. As most real-world data
do not come with an explicit schema, we develop a type inference algorithm that infers a type from 
representative sample documents and integrate it into the F\# type system using type providers.

Our library significantly reduces the amount of code that developers need to write when 
accessing data. It also provides additional safety guarantees, arguably, as much as possible if 
we abandon the closed world assumption.
\end{abstract}

@@

Social network clients, applications for finding tomorrow's weather or searching train schedules
all communicate with one or more services over the network and display the resulting information. 
Increasing number of such services provide REST-based end-points that return data as CSV, XML
or JSON. Despite numerous schematization efforts, most services do not come with an 
explicit schema. At best, the documentation provides sample responses for typical requests.

For example, OpenWeatherMap provides an end-point for getting current weather for a 
given city\footnote{See ``Current weather data'': \url{http://openweathermap.org/current}}. 
The documentation shows one sample to illustrate the typical response. Using standard 
JSON and web libraries, we might write:
%
\begin{equation*}
\begin{array}{l}
 \kvd{let}~\ident{doc}=\ident{Http.Request}(\str{http://api.owm.org/?q=NYC}) \\
 \kvd{match}~\ident{JsonValue.Parse}(\ident{doc})~\kvd{with} \\
 |~\ident{Record}(\ident{root})\rightarrow \\
 \quad \kvd{match}~\ident{Map.find}~\str{main}~\ident{root}~\kvd{with} \\
 \quad |~\ident{Record}(\ident{main})\rightarrow \\
 \quad \quad \kvd{match}~\ident{Map.find}~\str{temp}~\ident{main}~\kvd{with} \\
 \quad \quad |~\ident{Number}(\ident{num})\rightarrow \ident{printfn}~\str{Lovely \%f!}~\ident{num} \\
 \quad \quad |~\_\rightarrow \ident{failwith}~\str{Incorrect format} \\
 \quad |~\_\rightarrow \ident{failwith}~\str{Incorrect format} \\
 |~\_\rightarrow \ident{failwith}~\str{Incorrect format} 
\end{array}
\end{equation*}
%
The code assumes that the response has a particular format described in the documentation. The
root node must be a record with a \str{main} field, which has to be another record containing
a numerical \str{temp} field. When the format is incorrect, the data access simply fails
with an exception.

Using the JSON type provider from F\# Data, we can write code with exactly the 
same functionality in two lines:
%
\begin{equation*}
\begin{array}{l}
 \kvd{type}~\ident{W} = \ident{JsonProvider}\langl\str{http://api.owm.org/?q=NYC}\rangl \\[0.1em]
 \ident{printfn}~\str{Lovely \%f!}~(\ident{W.GetSample().Main.Temp})
\end{array}
\end{equation*}
%
$\ident{JsonProvider}\langl\str{...}\rangl$ invokes a type provider at 
compile-time with the URL as a sample. The type provider infers the structure of the response
and provides a type with a \ident{GetSample} method that returns a parsed JSON with nested
properties \ident{Main.Temp}, returning the temperature as a number.

The rest of the paper describes the mechanism and discusses its safety properties. 
The key novel contributions are:

@@

\item We describe a predictable type inference algorithm for structured data formats, 
  based on a \emph{common supertype} relation, that underlies the type providers 
  (\S\ref{sec:inference}).

\item We give a formal model (\S\ref{sec:formal}) and use it to prove
  \emph{relativized type safety} for the type providers (\S\ref{sec:safety}).
  This adapts the ML-style type safety for the context of the web.

\item The supplementary screencast show developer experience using 
  F\# Data with JSON, XML and CSV.
\end{itemize}

>>>>

\title{Types from data: \textnormal{Making structured data first-class citizens in F\#}}
%\subtitle{Subtitle Text, if any}

@@

In this paper, we integrate external structured data into the F\# type system. As most real-world data
does not come with an explicit schema, we develop a shape inference algorithm that infers a shape from 
representative sample documents and integrate it into the F\# type system using type providers.

Our library significantly reduces the amount of data access code and it provides additional 
safety guarantees when contrasted with the widely used weakly typed techniques.
\end{abstract}

@@

Social network clients, applications for finding tomorrow's weather or searching train schedules
all communicate with external services. Increasing number of such services provide REST-based 
end-points that return data as CSV, XML or JSON. Despite many schematization efforts, most 
services do not come with an explicit schema. At best, the documentation provides sample responses 
for typical requests.

For example, \url{http://openweathermap.org/current} provides an end-point to get the current weather.
The documentation shows one sample to illustrate the typical response. Using 
standard JSON and web libraries, we might write:

\noindent
\begin{equation*}
\begin{array}{l}
 \kvd{let}~\ident{doc}=\ident{Http.Request}(\str{http://api.owm.org/?q=NYC}) \\
 \kvd{match}~\ident{JsonValue.Parse}(\ident{doc})~\kvd{with} \\
 |~\ident{Record}(\ident{root})\rightarrow \\
 \quad \kvd{match}~\ident{Map.find}~\str{main}~\ident{root}~\kvd{with} \\
 \quad |~\ident{Record}(\ident{main})\rightarrow \\
 \quad \quad \kvd{match}~\ident{Map.find}~\str{temp}~\ident{main}~\kvd{with} \\
 \quad \quad |~\ident{Number}(\ident{num})\rightarrow \ident{printfn}~\str{Lovely \%f!}~\ident{num} \\
 \quad \quad |~\_\rightarrow \ident{failwith}~\str{Incorrect format} \\
 \quad |~\_\rightarrow \ident{failwith}~\str{Incorrect format} \\
 |~\_\rightarrow \ident{failwith}~\str{Incorrect format} 
\end{array}
\end{equation*}
%
The code assumes that the response has a particular shape described in the documentation. The
root node must be a record with a \str{main} field, which has to be another record containing
a numerical \str{temp} field. When the shape is different, the data access simply fails
with an exception. The code, while not immediately unsound, is manifestly prone to errors if
strings are misspelled or incorrect shape assumed.

Using the JSON type provider from F\# Data, we can write code with exactly the 
same functionality in two lines:
%
\vspace{-0.1em}
\begin{equation*}
\begin{array}{l}
 \kvd{type}~\ident{W} = \ident{JsonProvider}\langl\str{http://api.owm.org/?q=NYC}\rangl \\[0.1em]
 \ident{printfn}~\str{Lovely \%f!}~(\ident{W.GetSample().Main.Temp})
\end{array}
\end{equation*}
%
$\ident{JsonProvider}\langl\str{...}\rangl$ invokes a type provider at 
compile-time with the URL as a sample. The type provider infers the structure of the response
and provides a type with a \ident{GetSample} method that returns a parsed JSON with nested
properties \ident{Main.Temp}, returning the temperature as a number.

In short, \emph{the types come from the sample data}. In our experience, this technique is 
immensely practical and surprisingly effective in achieving sound information interchange 
in heterogeneous systems. This paper presents our approach:

@@

\item We describe a predictable shape inference algorithm for structured data formats, 
  based on a \emph{preferred shape} relation, that underlies the type providers 
  (\S\ref{sec:inference}).

\item We give a formal model (\S\ref{sec:formal}) and use it to prove
  \emph{relativized type safety} for the type providers (\S\ref{sec:safety}).
  This adapts the ML-style type safety for the context of the web.
\end{itemize}

\vspace{-0.1em}
\noindent
The supplementary screencast illustrates the practical developer experience using 
  F\# Data with JSON, XML and CSV.
